---
title: "Creating and transforming variables"
author: "Daniel_Asrat"
date: "2023-11-03"
output: prettydoc::html_pretty:
    theme: architect
    toc: true
    number_sections: false 
    editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: none
    
---
```{r warning=FALSE}
#  load the packages 
if(!require(pacman)) install.packages("pacman")   
pacman::p_load(rio, dplyr, janitor,tidyverse)

```

```{r}
# import data

obesity <- read_csv(here("C:/Users/danit/Downloads/obesity.csv"))

obesity

```

# removal of the first wto variable in the dataset

```{r}
obesity_new <- obesity %>%
  select(-1, -2)  # This selects all variables except the first two
obesity_new
```

# Clean data with {dplyr}

## **Step 1**: Examine and clean dataset (Instructor demo)

```{r, warning=FALSE}
obesity_clean <- obesity_new %>%
  mutate(
    # Convert these variables to numeric, NA will be introduced where conversion is not possible
    bmi = as.numeric(as.character(bmi)),  # Convert bmi to numeric
    sedentary_ap_s_day = as.numeric(as.character(sedentary_ap_s_day)),  # Convert sedentary_ap_s_day to numeric
    light_ap_s_day = as.numeric(as.character(light_ap_s_day)),  # Convert light_ap_s_day to numeric
    mvpa_s_day = as.numeric(as.character(mvpa_s_day)),  # Convert mvpa_s_day to numeric
    
     # Convert these variables to factors
    sex = as.factor(sex),  # Convert sex to factor
    status = as.factor(status),  # Convert status to factor
    oms_recommendation = as.factor(oms_recommendation)  # Convert oms_recommendation to factor
  ) 
```

# Checking the structure and summary of the dataset to ensure the changes are applied
```{r}
glimpse(obesity_clean)

summary(obesity_clean)

```

# Step 2: Convert physical activity variables to minutes per week and remove old variables

```{r}
# Step 2: Convert physical activity variables to minutes per week and remove old variables
obesity_clean1 <- obesity_clean %>%
  mutate(sedentary_ap_m_week = (sedentary_ap_s_day / 60) * 7,
    light_ap_m_week = (light_ap_s_day / 60) * 7,
    mvpa_m_week = (mvpa_s_day / 60) * 7) %>%
  select(-ends_with("_s_day"))  # Removes the old seconds per day variables

# Lastly, rename `oms_recommendation` to `who_recommendation`
obesity_clean2 <- obesity_clean1 %>%
rename(who_recommendation = oms_recommendation)
```


```{r}
# Check the changes with summary or glimpse
summary(obesity_clean2)
# or
glimpse(obesity_clean2)
```

## **Step 3**: Combine physical activity variables

```{r}
# Assuming `obesity_clean2` is your latest dataset from the previous step
obesity_clean3 <- obesity_clean2 %>% mutate(total_activity_m_week = light_ap_m_week + mvpa_m_week)
obesity_clean3
```

## **Step 4**: Replace values in selected columns

```{r}
# Step 4: Replace values in the 'sex' column
obesity_clean4 <- obesity_clean3 %>%
  mutate(sex = recode(sex, "M" = "Male", "F" = "Female"))

# Checking the changes
glimpse(obesity_clean4)
```
# Replace values in the 'status' column

```{r}
obesity_clean5 <- obesity_clean4 %>% mutate(status = recode(status, "Adulte" = "Adult", "Enfant" = "Child"))

# Checking the changes
glimpse(obesity_clean5)
```

# Visualize data with {esquisse} ## **Plot 1**: Histogram
```{r}
# esquisse::esquisser(obesity_clean5)

# Plotting a histogram of the bmi variable

ggplot(obesity_clean5, aes(x = bmi)) + 
  geom_histogram(binwidth = 1, color = "red", fill = "green") +
  labs(title = "Histogram of BMI",
       x = "BMI",
       y = "Frequency") +
  theme_minimal()
```

## **Plot 2**: Boxplot
```{r}
ggplot(obesity_clean5) +
  aes(x = bmi, y = sex, fill = sex) +
  geom_boxplot() +
  scale_fill_hue(direction = 1) +
  theme_minimal()
```
















